Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-26 manuscript.rmd
Manuscript draft 2 sent to co-authors

Section  | Word Count
------------- | -------------
Summary | 197
Main Text  | 6500
Introduction | 890
Methods | 3294
Results | 804
Discussion | 1512
Keywords | 6

@@

Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

# Results
## Soil water dynamics by depth
\
The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig. \@ref(fig:hydro-calib)A, B), including the out of sample observations. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)C) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Table S\@ref(tab:param)).

@@

# Conclusions
\
Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology. Trees realized water environments and water-access are nonetheless rarely studied. To our knowledge, this is the first study to demonstrate a mechanistic link between realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tree community. We report here that deep-water access plays a role in alleviating mortality of hydraulically vulnerable trees. This has important implications for our predictive understanding of tropical forest dynamics under current and future climate. Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.

@@

\
```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively. Only 100 curves are shown for clarity. Note the log scale on horizontal axis.")}
knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
```
\
```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs. TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
```
\
```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline). Growth rates are color coded by census interval. The model equation is shown in inset along with goodness-of-fit (Adj. $R$^2^).")}
knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model_no_color.jpeg"))
```
\
```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```
\
```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels). Species are color-coded by deciduousness.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
```
\
```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997. Data from @Meinzer:1999."}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```
\
```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines). Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
```
\
```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D). Each line represents a species, color coded by WSG (A, C), or LMA (B, D), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
```
\
```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates. Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
```
\
```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels). For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation. The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$. ERD models in panels A-E correspond to Eq. \\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```
\
```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD. $R$^2^ and $p$-values for linear model fits are given in panel insets. Model fits (blue lines) with SE (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
```

\
```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and slightly jittered horizontally for clarity.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_evergreen.jpeg"))
```

>>>>

Section  | Word Count
------------- | -------------
Summary | 197
Main Text  | 6516
Introduction | 890
Methods | 3300
Results | 814
Discussion | 1512
Keywords | 6

@@

\
Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

# Results
## Soil water dynamics by depth
\
The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig. \@ref(fig:hydro-calib)A, B), including the out of sample observations. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)C) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range (Table S\@ref(tab:param))  in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Fig. S\@ref(tab:sens-sm-et), S\@ref(tab:sens-run)).

@@

## Conclusions
\
Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology. Trees realized water environments and water-access are nonetheless rarely studied. To our knowledge, this is the first study to demonstrate a mechanistic link between realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tree community. We report here that deep-water access plays a role in alleviating mortality of hydraulically vulnerable trees. This has important implications for our predictive understanding of tropical forest dynamics under current and future climate. Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.

@@

```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively. Only 100 curves are shown for clarity. Note the log scale on horizontal axis.")}
knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
```

```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs. TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
```

```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline). Growth rates are color coded by census interval. The model equation is shown in inset along with goodness-of-fit (Adj. $R$^2^).")}
knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model_no_color.jpeg"))
```

```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```

```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels). Species are color-coded by deciduousness.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
```

```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997. Data from @Meinzer:1999."}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines). Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
```

```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Parameter sensitivity of ELM-FATES soil water content (VWC) is shown for two sample months. (A, B)) as well as for ET for another two sample months (C, D), as sensitivity for a particular parameter varies depending on the chosen month. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000-ensemble runs. See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
```

```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (A, B, C, D), as sensitivity for a particular parameter varies depending on the chosen month. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000-ensemble runs. See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
```

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D). Each line represents a species, color coded by WSG (A, C), or LMA (B, D), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
```

```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates. Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
```

```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels). For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation. The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$. ERD models in panels A-E correspond to Eq. \\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```

```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD. $R$^2^ and $p$-values for linear model fits are given in panel insets. Model fits (blue lines) with SE (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
```

```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and slightly jittered horizontally for clarity.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_evergreen.jpeg"))
```
